---
title: "ResourceSpectraFromESMOutput"
author: "phoebe.woodworth-jefcoats@noaa.gov"
output: html_document
date: "2024-10-11"
---

## How to use this script
This script includes a function to create mizer resource spectra from earth 
system model (ESM) output: `ESMspectrum`.  It also has some additional code and 
loads some data to walk you through an example.  

If you want to fully replicate the work here, all the necessary data are provided
in this folder.  If you have your own data ready to go, you can skim the script
and jump to the `ESMspectrum` function.  And, if you have questions or need a hand, please
[open an issue](https://github.com/pwoodworth-jefcoats/therMizer-WCPFC/issues). 
Someone else probably has the same questions you do, so you'll be helping people 
out by doing this.

We'll get the `ESMspectrum` function linked to the [mizer](https://github.com/sizespectrum/mizer)
or [therMizer](https://github.com/sizespectrum/therMizer) packages soon!  
Once they're there, you can use them without this guide.  In the meantime, you 
could copy and paste just the function for use in your own modeling.  

### An example for the Hawaiʻi-based deep-set longline fishery
This example uses a model configured for the Hawaiʻi-based deep-set longline 
fishery.  You can read more about the system in [Woodworth-Jefcoats et al. 2019](https://www.frontiersin.org/journals/marine-science/articles/10.3389/fmars.2019.00383/full).
You can find code for accessing and packaging ESM output in a [repository](https://github.com/pwoodworth-jefcoats/therMizer-FishMIP-2022-HI)
which documents a FishMIP contribution for this system.  This script assumes you
have ESM output ready to go, but if you have questions about how to tackle that
piece, you can [open an issue](https://github.com/pwoodworth-jefcoats/therMizer-WCPFC/issues).

#### Data provided
The following data are included in this folder and used to run this script:  
* HIregion_species_params.csv: species parameters for this model  
* HIregion_inter.csv: species interaction matrix for this model  
* GFDLplankton.csv: matrix with ESM-output plankton carbon densities (more on
this below)  
* GFDLsizes.csv: size class midpoints for ESM plankton groups (more on this below)

#### Set up your mizer model
Here, we'll set up the multi-species mizer model that we want to force with the 
ESM output.  **Note:** You need to do this step before you can use the 
`ESMspectrum` function, which uses the `w_full` parameter.  

Okay, let's get started with the necessary and helpful packages:
```{r, message = FALSE}
# Set up environment
library(mizer) # to run mizer
library(here) # handy for directory navigation
```

Now, we'll set up a very basic multi-species model:
```{r}
# Load parameters 
HIparams <- read.csv("HIregion_species_params.csv", header = TRUE)
HIinter <- read.csv("HIregion_inter.csv", row.names = 1)
HIinter <- as(HIinter,"matrix")

# Generate parameters
params <- newMultispeciesParams(HIparams, interaction = HIinter, kappa = 1e12, min_w_pp = 1e-14, w_pp_cutoff = 455400) 
```
Now that that's done, we can move on to working with the ESM output. 

#### Create resource spectra from ESM output
Now, we're ready to use the ESM output to create resource spectra.  Exciting!
This example uses output from the GFDL-MOM6-COBALT2 model that was used in 
FishMIP's 2022 ISIMIP 3a protocol.  You can use whatever ESM you want.  You just
need:  
* Total plankton carbon (in grams) for each size class for each time step of your
simulation.  This is what's provided in `GFDLplankton.csv`.   
* Values for the mid-point of each size class (in grams wet weight).  This is what's 
provided in `GFDLsizes.csv`.  
* Values for sizes of your background resource.  These should
be from mizer's `w_full` parameter.  This example uses the full size range because
the simulation focuses on top predators, but you might want a smaller size range.  

If you need to convert plankton size classes from Equivalent Spherical Diameter (ESD)
in $\mu$m to grams wet weight, you can use this equation:  
$\frac{4}{3}\pi(0.5\times0.0001\times size)^3$
**NOTE TO SELF: add citation for this**
If you'd like this functionality added to the `ESMspectrum` function, simply
[open an issue](https://github.com/pwoodworth-jefcoats/therMizer-WCPFC/issues) to
request it. 

Determining the size ranges of ESM-output plankton groups requires some sleuthing.
The best advice I can offer is to dig into the model's documentation as best you
can.  I also recommend [this handy paper](https://www.frontiersin.org/journals/marine-science/articles/10.3389/fmars.2021.622206/full) 
by Kearney et al. 2021 which should help you get started (full disclosure: I am 
a coauthor of this paper).

Okay, let's get started with loading the ESM-output:
```{r}
GFDLplankton <- read.csv(here("PlanktonFunction", "GFDLplankton.csv"), header = TRUE)
```
`GFDLplankton` is a matrix of plankton densities (in grams carbon) for each 
size class (columns) for each time step (rows).  I've arranged them from smallest
to largest, which is what I'd recommend.  Column names are the size classes.

Let's also load the size class mid-points:
```{r}
GFDLsizes <- read.csv(here("PlanktonFunction", "GFDLsizes.csv"), header = TRUE)
```
`GFDLsizes` is a matrix of plankton size class mid-points (in grams wet weight).
The column headers match in name and order `GFDLplankton`.  

This is all you need, along with `w_full`, to use the function below to convert 
the ESM output into resource spectra.  Here's what the function does:  
* Converts plankton biomass densities from grams carbon to grams wet weight by 
multiplying by 10,  
* Converts plankton biomass densities to numerical abundance by dividing each 
size class density by its mid-point size.  This step assumes that all plankton 
are the mid-point size.  Obviously this isn't the case, but it keeps the math
straightforward.  I'm game to experiment with additional approaches.  If you have
an idea and want to collaborate, please [open an issue](https://github.com/pwoodworth-jefcoats/therMizer-WCPFC/issues).
Alternatively, you can repurpose this code to suit your needs.  
* Creates a linear model of numerical abundance versus size at each time step and
uses the slope and intercept of that model to create a resource spectrum over 
the size range you specify.  The default is the full range of `w_full`, but you
can specify something different.  
* Includes the options to set values by which to scale the linear model's slope
and intercept.  This is often needed in order to generate resource spectra that
allow your model to stabilize.  ESMs aren't designed to fully replicate real plankton
communities, which is why this scaling is often necessary.
* Returns 3 things:  
  + Background resource array with the dimensions of time $\times$ size  
  + The slope from the linear model at each time step, for diagnostic purposes
  + The intercept from the linear model at each time step, for diagnostic purposes
  
Here's the function:
```{r}
ESMspectrum <- function(plankton_array, 
                        plankton_sizes, 
                        resource_sizes = params@w_full, 
                        slope_scale = 1, 
                        intercept_scale = 1){
  # Numerical abundance from biomass densities converted to gww
  plankton_abundance <- plankton_array*10/as.vector(plankton_sizes)
  
  # Size classes for linear model (in log space)
  plankton_x <- as.numeric(log10(plankton_sizes))
  
  # Size classes for resource spectra, again in log space
  resource_x <- log10(resource_sizes)
  
  # Use linear model to generate resource spectra at each time step
  # Identify number of time steps
  t_max <- dim(plankton_array)[1]
    # Identify number of size classes
  sizes <- length(resource_x)
  
  # Create array and fill it
  resource_spectra <- array(numeric(), c(t_max,sizes)) # time by size
  resource_slope <- array(numeric(), c(t_max,1)) 
  resource_intercept <- array(numeric(), c(t_max,1)) 
  
  # y values
  for (t in seq(1,t_max,1)) {
    resource_plankton <- as.numeric(log10(plankton_array[t,]))
    
    # Calculate slope and intercept, expand spectra for full size range
    # Linear fits
    resource_lm <- lm(resource_plankton ~ plankton_x)
    
    # Expand to full size range
    resource_spectra[t,] <- resource_lm$coefficients[2] * resource_x * slope_scale + (resource_lm$coefficients[1] * intercept_scale)
    
    # Save slope and intercept, for diagnostics
    resource_intercept[t,1] <- resource_lm$coefficients[1]
    resource_slope[t,1] <- resource_lm$coefficients[2]
  }
  
  return(list("resource_spectra" = resource_spectra, 
              "resource_slope" = resource_slope, 
              "resource_intercept" = resource_intercept))
}
```

And here are some examples of how to use it:
```{r}
Resource <- ESMspectrum(GFDLplankton, GFDLsizes)
```

Yet to add:  
* Attaching the resource to the parameters  
* Examples with scaling, including why you'd scale the resource  
* Running mizer with the ESM resource  
* Additional explanatory text as helpful  

