---
title: "ResourceSpectraFromSatelliteData"
author: "phoebe.woodworth-jefcoats@noaa.gov"
output: html_document
---

## How to use this script
This script includes a function to create mizer resource spectra from satellite
remotely sensed data: `SATspectrum`.  It also includes a function, 
`plankton_forcing`, for forcing your mizer model with the derived resource
spectra.  And, it has some additional code and loads some data to walk you 
through an example.  

If you want to fully replicate the work here, all the necessary data are provided
in this folder.  If you have your own data ready to go, you can skim the script
and jump to the `SATspectrum` and `plankton_forcing` functions.  And, if you have 
questions or need a hand, please [open an issue](https://github.com/pwoodworth-jefcoats/therMizer-WCPFC/issues). 
Someone else probably has the same questions you do, so you'll be helping people 
out by doing this.

We'll get the `SATspectrum` function linked to the [mizer](https://github.com/sizespectrum/mizer)
or [therMizer](https://github.com/sizespectrum/therMizer) packages soon! 
Once they're there, you can use them without this guide.  In the meantime, you 
could copy and paste just the functions for use in your own modeling.  

### An example for the oligotrophic North Pacific 
This example uses a model configured for the Hawaiʻi-based deep-set longline 
fishery, whose operating footprint includes the oligotrphic North Pacific.  A 
small piece of ocean is being used here in the interest of expediency as a
proof-of-concept.  You can read more about the system in [Woodworth-Jefcoats et al. 2019](https://www.frontiersin.org/journals/marine-science/articles/10.3389/fmars.2019.00383/full).  

This script assumes you have satellite data ready to go.  But, if you have 
questions about how to tackle this piece, there's code in the [PlanktonSize](https://github.com/pwoodworth-jefcoats/therMizer-WCPFC/tree/main/PlanktonSize)
folder of this repository.  You can also [open an issue](https://github.com/pwoodworth-jefcoats/therMizer-WCPFC/issues)
with any questions.

#### Data provided
The following data are included in this folder and used to run this script:  

* `HIregion_species_params.csv`: species parameters for this model  
* `HIregion_inter.csv`: species interaction matrix for this model  
* `Resource_Min.csv`: smallest phytoplankton size, in g C  
* `Resource_Max.csv`: largest phytoplankton size, in g C  
* `Resource_Slope.csv`: slope of the plankton spectrum  
* `Resource_Intercept.csv`: intercept of the plankton spectrum

For now, the resource data are just from a single time step.  Eventually, these 
could be time series as are used the the `ESMspectrum` function.

#### Set up your mizer model
Here, we'll set up the multi-species mizer model that we want to force with the 
ESM output.  **Note:** You need to do this step before you can use the 
`SATspectrum` function, which uses the `w_full` parameter.  

Okay, let's get started with the necessary and helpful packages:
```{r, message = FALSE}
# Set up environment
library(mizer) # to run mizer
library(here) # handy for directory navigation
```

Now, we'll set up a very basic multi-species model:
```{r}
# Load parameters 
HIparams <- read.csv("HIregion_species_params.csv", header = TRUE)
HIinter <- read.csv("HIregion_inter.csv", row.names = 1)
HIinter <- as(HIinter,"matrix")

# Load plankton size bounds
phyto_min <- read.csv("Resource_Min.csv")
phyto_max <- read.csv("Resource_Max.csv")

# Generate parameters
# params <- newMultispeciesParams(HIparams, interaction = HIinter, kappa = 1e12, min_w_pp = 1e-14, w_pp_cutoff = 455400) 
params <- newMultispeciesParams(HIparams, interaction = HIinter, kappa = 1e12, 
                                min_w_pp = phyto_min$plankton_min[1], 
                                w_pp_cutoff = 455400) 
```
Now that that's done, we can incorporate the plankton slope and intercept. 

#### Create resource spectra from ESM output
Now, we're ready to use the satellite data to create resource spectra.  Exciting!
This example uses [sea surface temperature](https://oceanwatch.pifsc.noaa.gov/erddap/griddap/CRW_sst_v3_1_monthly.html) 
and [ocean color](https://oceanwatch.pifsc.noaa.gov/erddap/griddap/esa-cci-chla-monthly-v6-0.html) 
data from [OceanWatch](https://oceanwatch.pifsc.noaa.gov/).  You can use other 
sources if you prefer.  

This example follows the methodology of [Barnes et al. (2011)](https://academic.oup.com/plankt/article/33/1/13/1403550.
See the [PlanktonSize](https://github.com/pwoodworth-jefcoats/therMizer-WCPFC/tree/main/PlanktonSize)
folder of this repository for details on its implementation.  

Okay, let's get started by loading in the spectrum slope and intercept:
```{r}
phyto_slope <- read.csv("Resource_Slope.csv")
phyto_intercept <- read.csv("Resource_Intercept.csv")
```

We'll also replicate these values for, say, 50 time steps just so the model has
something to work with:
```{r}
phyto_slope_ts <- rep(phyto_slope$slope[1], 50)
phyto_intercept_ts <- rep(phyto_intercept$intercept[1], 50)
```

This is all you need, along with `w_full`, to use the function below to convert 
the satellite remotely sensed data into resource spectra.  Here's what the function does:  

* Includes the options to set values by which to scale the linear model's slope
and intercept.  This is often needed in order to generate resource spectra that
allow your model to stabilize.  Creating a single apprpach that accurately 
captures all ocean waters is a tall order, which is why this scaling may be
necessary.  
* Includes the option to specify time steps if you want to use something specific
in your mizer model.  The default is time steps numbered 0 through the maximum number
of time steps in your plankton data.
* Returns 2 things:  
  + Background resource array with the dimensions of time $\times$ size  
  + Background resource array ready as `n_pp_array` for attaching to your model  

Here's the `SATspectrum` function:
```{r}
SATspectrum <- function(plankton_slope, 
                        plankton_intercept, 
                        resource_sizes = params@w_full, 
                        slope_scale = 1, 
                        intercept_scale = 1,
                        time_steps = seq(0, length(plankton_slope)[1], 1)){
  
  # Size classes for resource spectra, in log space
  resource_x <- log10(resource_sizes)
  
  # Use linear model to generate resource spectra at each time step
  # Identify number of time steps
  t_max <- length(plankton_slope)
  # Identify number of size classes
  n_sizes <- length(resource_x)
  
  # Create array and fill it
  resource_spectra <- array(numeric(), c(t_max,n_sizes)) # time by size
  
  # y values
  for (t in seq(1,t_max,1)) {
    # Apply linear model
    resource_spectra[t,] <- plankton_slope[t] * resource_x * slope_scale + (plankton_intercept[t] * intercept_scale)
  }
  
  # Create n_pp_array
  sizes <- names(params@initial_n_pp)
  n_pp_array <- array(NA, dim = c(length(time_steps), length(sizes)), dimnames = list(time = time_steps, w = sizes))
  
  # Fill n_pp_array
  n_pp_array[1,] <- resource_spectra[1,]
  for (t in seq(1,length(time_steps) - 1,1)) {
    n_pp_array[t + 1,] <- resource_spectra[t,]
  }
  
  return(list("resource_spectra" = resource_spectra, 
              "n_pp_array" = n_pp_array))
}
```

And here's how to use it in its simplest implementation:
```{r}
Resource <- SATspectrum(phyto_slope_ts, phyto_intercept_ts)
```

#### Use the resource spectra built from satellite data
But we want to do more than generate the resource spectra.  We want to use them 
to force a mizer model.  To do that, we need to attach the spectra to the model
parameters and tell mizer to use it.  This is where the `plankton_forcing`
function comes in:
```{r}
# From therMizer, create a plankton_forcing function
plankton_forcing <- function(params, t, ...) {
  w_cut_off <- params@resource_params$w_pp_cutoff
  # index selects the right resource for any t
  index_vec <- as.numeric(dimnames(other_params(params)$n_pp_array)[[1]]) - t
  index <- which(index_vec == max(index_vec[index_vec<=0]))
  # Convert to density
  pkt <- 10^(other_params(params)$n_pp_array[index,])/params@dw_full 
  # Set sizes above plankton resource max to zero
  pkt[which(as.numeric(names(pkt)) >= w_cut_off)] <- 0
  return(pkt)
}
```

Then, we need to tell mizer to use it and the new resource spectra (`n_pp_array`):
```{r}
# Also from therMizer, attach the n_pp_array using this function
params <- setResource(params, resource_dynamics = "plankton_forcing")

# Attach n_pp_array to params
other_params(params)$n_pp_array <- Resource$n_pp_array
```

And, finally, use it in a very simple case without fishing:
```{r}
InitialSim <- project(params, dt = 1/12, t_max = 50, t_save = 1, effort = 0)
plot(InitialSim)
```

#### Scaling the background resource
The initial test we did above demonstrates why you'd want to scale the background 
resource.  Earth system models aren't designed to model plankton communities, so
their output a) doesn't always resemble real ecosystems, and b) isn't often 
directly comparable across models.  Scaling the ESM-derived resource spectra
can resolve both these issues.  An alternative would be to bias correct in the 
way that's often done with temperature.  If you have observations you can use for
this, go for it.  Most systems, including the one modeled here, lack those data.
Hence, scaling.

In order to have something to "ground truth" your scaling attempts to, you'll want
to run mizer with your species and system parameters, and with mizer's built-in
semi-chemostat resource.  Use the feeding levels and biomass densities that 
result from that to guide your scaling efforts.  This is admittedly a bit tedious
and requires iterative trial and error.  I'm open to alternatives.  Please 
[open an issue](https://github.com/pwoodworth-jefcoats/therMizer-WCPFC/issues) to 
suggest them. 

For the purposes of this example, I'm going to draw on the earlier work linked 
above and scale the spectra slope by 1.03 and the intercept by 0.85:
```{r}
ScaledResource <- ESMspectrum(GFDLplankton, GFDLsizes, slope_scale = 1.03, intercept_scale = 0.85)
```

Attach this new resource to a new set of parameters:
```{r}
new_params <- newMultispeciesParams(HIparams, interaction = HIinter, kappa = 1e12, min_w_pp = 1e-14, w_pp_cutoff = 455400)
new_params <- setResource(new_params, resource_dynamics = "plankton_forcing")
other_params(new_params)$n_pp_array <- ScaledResource$n_pp_array
```
And run another simple simulation without fishing: 
```{r}
ScaledSim <- project(new_params, dt = 1/12, t_max = 600, t_save = 10, effort = 0)
plot(ScaledSim)
```



Yet to add:  

* Additional explanatory text as helpful  
* Checks to make sure the right sizes classes are being applied to each other
* Other stuff you'd like to see? 

## References
Barnes C, Irigoien X, De Oliveira JAA, Maxwell D, and Jennings S. 2011. 
Predicting marine phytoplankton community size structure from empirical 
relationships with remotely sensed variables, Journal of Plankton Research. 
33(1):13–24, https://doi.org/10.1093/plankt/fbq088

Woodworth-Jefcoats PA, Blanchard JL and Drazen JC. 2019. Relative impacts of 
simultaneous stressors on a pelagic marine ecosystem. Front. Mar. Sci. 6:383.
doi: 10.3389/fmars.2019.00383